Off the beaten path: \\ Gaia reveals GD-1 stars outside of the main stream

### 2018-07-24 paper/submit-aas/referee-response2.txt

> However, the maps sent to the referee but not included in the paper do
> not ease my concerns on how robust their conclusions are. For example,
> the figures show that the assumption of a uniform density may be
> flawed given the number of features in the data. Especially the region
> around the spur shows a high density of large values of the unit
> weight error.
>
> Since their procedure starts with a selection of tentative members
> based on proper motions, it would be good if the authors could make a
> figure that shows both the stars selected according to the proper
> motion (i.e. the sample they already have) and the stars with 2
> parameter solutions that fall within the CMD selection. It would also
> be useful to assess the relative number of good and bad solutions in
> the blob and the spur, as well as in the regions around the gaps.
>
> The issue is not so much the quality of the data that is kept (the 5
> parameter solutions), but what does incompleteness in this dataset
> imply? The presence of stars with only 2 parameter solutions implies
> that such stars would not be selected according to the procedure
> followed by the authors, and their exclusion could potentially lead to
> "features". A plot as I propose would help to double check that the
> exclusion of such stars (if any) does not give rise to gaps or
> spurious over-densities.

We have attached the requested figure to the report. We do not see correlations between the off-track features described in this paper and the CMD-selected stars with 2-parameter solutions.

@@

> I would also like to ask the authors to overplot their selected GD-1
> stream stars (i.e. those shown in the bottom panel of Fig. 1), in the
> maps of the density of 2 parameter solutions, the unit weight
> error and the extinction map.

We have attached these figures to the response. Again, while there is clearly structure in these maps, we do not see any relation between the data quality or incompleteness and the new off-track features reported in this paper. We do see issues with completeness at phi1 < -50 due to the scan pattern, and we discuss this in the text.

> Finally a small point regarding the presence of a progenitor. Given
> that it is very difficult to identify the feature the authors
> associated to the progenitor clearly from the figures presented, and
> its significance is not demonstrated in the paper, it does not really
> seem to deserve a mention in the abstract.

We agree that the uniqueness of the previous claim was not strong, and we have both rephrased the text and added an alternate model stream for comparison. With Figure 2, we now show two plausible locations for the progenitor: one in which the progenitor disrupts at present day and results in the over-density, and one in which the progenitor disrupted several orbits ago and can explain one of the observed under-densities.

### 2018-07-24 paper/gd1.tex
toy model 2 rephrases

% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material; we defer a robust analysis of variations in the
stream width to future work.
\changes{With present data, however, we cannot rule out a scenario in which the progenitor system disrupted longer ago and resulted in one of the larger gaps (e.g., \citealt{DeBoer:2018}).}
% that has not yet phase-mixed in multiple dimensions.

@@

\changes{To initialize the model stream generation, we assume that the present day position of the progenitor system is either at $\phi_1 = -13.5^\circ$, the peak of the observed surface density profile (toy model 1), or was fully-disrupted 500 Myr in the past and the under-density at $\phi_1 = -20^\circ$ represents the fully-disrupted progenitor (toy model 2).}
We numerically integrate backwards from these locations for $4~\textrm{Gyr}$,
then forward-generate the model streams, releasing stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until its full disruption at the \changes{present (toy model 1), or at $t = -500~\textrm{Myr}$ (toy model 2)}.
% day, where we assume $M = 0~\msun$
% (i.e. fully disrupted).
Stream properties computed from the resulting model streams are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange and purple) lines.
The model streams qualitatively match the stream track in all phase-space
dimensions, but have much smoother surface density profiles along the extent of
the stream.
The broadening in the observed stream at negative longitudes is also seen in the model stream and is expected for older parts of the stream that have phase-mixed for longer times.
We therefore don't expect that the observed surface density variations can be fully explained by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).

>>>>

% First, the visible portions of the GD-1 stream are located at high Galactic
% latitudes ($b > 20^\circ$).
%, and we therefore do not expect significant dust
% extinction or variations in extinction within the stream region.
% The median $V$-band extinction over the stream track is
% $\textrm{med}\left(A_V\right) = 0.04~\textrm{mag}$ with a dispersion
% $\sigma_{A_V} \approx 0.03~\textrm{mag}$, as computed from the SFD extinction
% map (\citealt{Schlegel:1998}).
% As the stream approaches the Galactic disk ($\phi_1 < -60^\circ$),
% dust extinction becomes appreciable, and the stream stars are harder to
% select apart from the background.
% Second, 99\% of the stars in our sample have at least 8 \gaia\ visibility
% periods (median and root-variance of 14 and 3, respectively), promising a robust
% estimate of the proper motions.
% The variance in the number density of faint stars originating from the \gaia\
% scanning pattern is only $\approx 1~\textrm{deg}^{-2}$, as estimated using stars
% with proper motions $-9 < \mu_{\phi_1, \odot} < -4.5~\masyr$ and $1 <
% \mu_{\phi_2, \odot} < 4~\masyr$, far smaller than the density variations in the
% stream.
% The relative ratios between different over- and under-densities might be affected by the \gaia\ completeness at the faint end, but they
The observed density variations are therefore likely real morphological changes
along the GD-1 stream.
% as have been suggested before photometric selection alone
% (\citealt{Carlberg:2013, DeBoer:2018}).
The deep under-densities we see from the kinematically-cleaned sample correspond
to previously reported gaps in the stream (\citealt{Carlberg:2013, DeBoer:2018}).
There is also a well-defined, sharp overdensity in the surface density profile
close to $\phi_1 \approx -13^\circ$ with roughly symmetric under-densities on
either side:
this is suggestive of a progenitor system in the final stages of dissolution
(\citealt{Balbinot:2018}).
This region is also the thinnest part of the stream, as would be expected for
recently stripped material; we defer a robust analysis of variations in the
stream width to future work.
\changes{With present data, however, we cannot rule out a scenario in which the progenitor system disrupted longer ago and resulted in a gap at $\phi_1\approx-20^\circ$ (see Figure~\ref{fig:track-and-model} and, e.g., \citealt{DeBoer:2018}).}
% that has not yet phase-mixed in multiple dimensions.

@@

\changes{We present two progenitor configurations which reproduce certain aspects of the observed GD-1 density variations: the progenitor system is either at $\phi_1 = -13.5^\circ$, the peak of the observed surface density profile (toy model 1), or was fully-disrupted 500\,Myr in the past and the under-density at $\phi_1 = -20^\circ$ represents the fully-disrupted progenitor (toy model 2).}
% \changes{To initialize the model stream generation, we assume that the present day position of the progenitor system is either at $\phi_1 = -13.5^\circ$, the peak of the observed surface density profile (toy model 1), or was fully-disrupted 500 Myr in the past and the under-density at $\phi_1 = -20^\circ$ represents the fully-disrupted progenitor (toy model 2).}
We numerically integrate backwards from these locations for $4~\textrm{Gyr}$,
then forward-generate the model streams, releasing stream particles at each
timestep.
We assume an initial progenitor mass of $M=10^5~\msun$, and linearly decrease
the mass of the progenitor until its full disruption at the \changes{present (toy model 1), or at $t = -500~\textrm{Myr}$ (toy model 2)}.
% day, where we assume $M = 0~\msun$
% (i.e. fully disrupted).
Stream properties computed from the resulting model streams are plotted in
\figurename~\ref{fig:track-and-model} as dashed (orange and purple) lines.
The model streams qualitatively match the stream track in all phase-space
dimensions, but have much smoother surface density profiles along the extent of
the stream.
The broadening in the observed stream at negative longitudes is also seen in the model stream and is expected for older parts of the stream that have phase-mixed for longer times.
We therefore don't expect that the observed surface density variations can be fully explained by, e.g., epicyclic overdensities (\citealt{Kupper:2012}).
